Type Classes for Mathematics in Type Theory

### 2010-06-28 jar.tex
Revise abstract to make it more in line with title. Fix typo.

\begin{document}
\title{Type Classes for Mathematics in Type Theory}
\author{Bas Spitters \and Eelis van der Weegen} % For MSCS \\ Radboud University Nijmegen}
\institute{Radboud University Nijmegen}
\date{today}
\maketitle
\begin{abstract}
We present a new formalization of the algebraic hierarchy in Coq, exploiting its new type class mechanism to make practical a solution formerly thought infeasible. Our approach addresses both traditional challenges as well as new ones resulting from our ambition to build upon this development a library of constructive analysis in which abstraction penalties inhibiting efficient computation are reduced to a minimum. To support mathematically sound abstract interfaces for $\N$, $\Z$, and $\Q$, our formalization includes portions of category theory and multisorted universal algebra.
Algebra flourishes by the interplay between syntax and semantics. The Prolog-like abilities of type class unification allow us to conveniently define a quote function thus facilitating the use of reflective techniques.\marginpar{Do TODOs}
  % todo: handle.
\end{abstract}

@@

Some theory and programs about naturals are more easily proved, respectively defined, for concrete implementations and then ``lifted'' to the abstract interface. For relatively simple things like decidability of equality, all that is needed to make this lifting easy is a handful of simple lemmas translating the initiality property back into statements saying that \lstinline|naturals_to_semiring| is extensionally equal to other semiring morphisms and involutive between \lstinline|Naturals| implementations. For the example of decidability, these let us define decidability for an arbitrary \lstinline|Naturals| implementation simply by mapping to the \lstinline|nat| implementation using \lstinline|naturals_to_semiring|, deciding equality there, and reflecting the equality or inequality proof back to the abstract implementation using involutivity of \lstinline|naturals_to_semiring|.\marginpar{Word: Involutive?}

>>>>

\begin{document}
\title{Type Classes for Mathematics in Type Theory}
\author{Bas Spitters \and Eelis van der Weegen} % For MSCS \\ Radboud University Nijmegen}
\institute{Radboud University Nijmegen}
\date{today}
\maketitle
\begin{abstract}
The introduction of first-class type classes in the Coq system calls for re-examination of the basic interfaces we use for mathematical formalization in type theory.
% especially since existing solutions fall short from a proof engineering perspective.
We present a new set of foundational type classes for mathematics that take full advantage of their unique features to make practical a particularly flexible approach formerly thought infeasible. Thus, we address both traditional proof engineering challenges as well as new ones resulting from our ambition to build upon this development a library of constructive analysis in which abstraction penalties inhibiting efficient computation are reduced to a minimum.

The base of our development consists of type classes representing a standard algebraic hierarchy, as well as portions of category theory and universal algebra. We build upon this foundation a set of mathematically sound abstract interfaces for different kinds of numbers, succinctly expressed using categorical language and universal algebra constructions. Strategic use of type classes lets us support these high-level theory-friendly definitions while still enabling efficient implementations unhindered by gratuitous indirection or conversion or projection.

Algebra flourishes by the interplay between syntax and semantics. The Prolog-like abilities of type class instance resolution allow us to conveniently define a quote function, thus facilitating the use of reflective techniques.
\end{abstract}

@@

Some theory and programs about naturals are more easily proved, respectively defined, for concrete implementations and then ``lifted'' to the abstract interface. For relatively simple things like decidability of equality, all that is needed to make this lifting easy is a handful of simple lemmas translating the initiality property back into statements saying that \lstinline|naturals_to_semiring| is extensionally equal to other semiring morphisms and involutary between \lstinline|Naturals| implementations. For the example of decidability, these let us define decidability for an arbitrary \lstinline|Naturals| implementation simply by mapping to the \lstinline|nat| implementation using \lstinline|naturals_to_semiring|, deciding equality there, and reflecting the equality or inequality proof back to the abstract implementation using involutivity of \lstinline|naturals_to_semiring|.\marginpar{Word: Involutive?}
